Relationship between solar wind, $D_{st}$, and plasmasphere mass density on one-hour time scales

### 2015-09-03 paper.tex
Paper edits

\cite{Takahashi2010} developed a mass density dataset using measurements from the GOES satellites over the years 1980 through 1992, with most measurements in the range $L=6.8\pm0.2$. The equatorial mass density, $\rho_{eq}$ was estimated from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}(LR_E/R)$. The mass density $\rho$ was estimated using the the Alfven wave velocity relationship, $V_A=(\mu_0\rho)^{1/2}$, a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of an infinitely conducting ionosphere.

@@

In this work, events defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated in by horizontal dashed lines in the respective panels of Figure~\ref{AllData}.  In finding events, all fill values were replaced with linearly interpolated values.  Figure \ref{AllData} shows the values of $B_z$, $V_{sw}$, $D_{st}$, $F_{10.7}$, and $\rho_{eq}$ for the duration of the dataset.

@@

Figure~\ref{ccplot:a} shows the long-term trends of $log(\rho_{eq})$ and $F_{10.7}$ computed using the median values in 27-day (non-overalpping?) windows using all non--fill hourly values.  A scatter plot of these two lines is shown in Figure~\ref{ccplot:b}.  The linear correlation is found to be $0.87$ (? plot says 0.94!).  In comparison, \cite{Takahashi2010} found a correlation coefficient of $0.94$ using measurements from all satellites in the time interval of 1980 through 1991.

\subsection{D$_{st}$ and $\rho_{eq}$ Events}

@@

In the upper panel of \label{DailyAverages}, $D_{st}$ events are shown.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is ~2 amu/cm$^3$ instead of the value of ~10 amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for $D_{st}$, the number of measurements used in computing the medians is much larger as we use all available $D_{st}$ measurements instead of restricting to only values where a non--fill $\rho_{eq}$ value existed.

In the lower panel of \label{DailyAverages}, $\rho_{eq}$ events are shown.  Here we see that $\rho_{eq}$ enhancements do not clearly correspond to a signature in $D_{st}$, but that there is a trend for such enhancments to occur when $B_{z}$ in the IMF is positive.  This is consistent with \cite{Takahashi2010} who note that very large increases in $\rho_{eq}$ often corresponded to movement of the plasmasphere to beyond geosynchronous distances.

This method finds 669 such periods between May 1983 and August 1991 with an average duration of 9 hours and a median duration of 3 hours. Figure \ref{Storms} shows the average values of all events over a window of 24 hours before onset and 48 hours after. Figure \ref{Storms}a shows events selected by looking for event onsets where $D_{st}$ crossed a threshold of $-40nT$. The dashed red lines indicate plus and minus one standard deviation of values from all events that went into the average. The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack. A subset of longer duration events will be looked at later.

This figure shows a definite spike in the $Z$ component of the magnetic field, as well as the defined drop in $DST$, but no obvious change in mass density at an hourly timescale. This points to an issue with only looking at long-timescale trends between density and $D_{st}$, and allows for the possibility that other factors are influencing the long term correlation since there's no obvious connection on a short timescale. One possibility is that, as suggested in \cite{Takahashi2010}, $F_{10.7}$ plays a significant role in driving long term density values which biases the long term correlation of density and $D_{st}$.

@@

In the upper panel of \label{DailyAverages}, $D_{st}$ events are shown.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events corrspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is on the order of $\sim2$~amu/cm$^3$ instead of $\sim10$~amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for $D_{st}$, the number of measurements used in computing the medians is much larger as we use all available $D_{st}$ measurements instead of restricting to only values where a non-fill $\rho_{eq}$ value existed.

In the lower panel of \label{DailyAverages}, $\rho_{eq}$ events are shown.  Here we see that $\rho_{eq}$ enhancements do not clearly correspond to a signature in $D_{st}$, but that there is a trend for such enhancments to occur when $B_{z}$ in the IMF is positive.  This is consistent with \cite{Takahashi2010} who note that very large increases in $\rho_{eq}$ often corresponded to movement of the plasmasphere to beyond geosynchronous distances.

@@

\section{More specific events}
It's hypothesized that progressively picking more specific event criteria will allow for the possibility of more significant results, at the expense of more bias in the selection process and potentially less overall usefulness of the results. That said, the predictability of extreme events is of definite interest, so an attempt has been made to find some reproducible method of prediction. Looking at events that last longer than 12 hours and events with an onset threshhold greater than $70g/cm^3$ results in the left and right sides of Figure \ref{Mspec} respectively.

Neither of these seem to indicate anything too significant, so looking at $D_{st}$ events instead to look for something that causes a significant change in Mass Density results in Figure \ref{Dspec}. This shows that by either looking only at $D_{st}$ events that last longer than an hour (left) or at events where the onset condition is $D_{st}<-80nT$, a spike in mass density is seen, but also a definite lack of data availability to the point where that spike may be coming from less than five of the total 143 events.

Unfortunately there are no events in this time frame that are longer than 12 hours with $D_{st}$ minima lower than $-80nT$ that have existing mass density data around onset, so an analysis of this particular relationship can't be made.

@@

\begin{figure}[htp!]
\centering
\subcaptionbox{\label{ccplot:a}}
{\includegraphics[scale=0.40]{paperfigures/F107MDAllData.eps}}
\subcaptionbox{\label{ccplot:b}}
{\includegraphics[scale=0.40]{paperfigures/ccplot.eps}}
\caption{(a) 27-day averages of $F_{10.7}$ and $\rho_{eq}$. (b) Correlation between $log(\rho_{eq})$ and $F_{10.7}$ using hour, day, and 27-day averaging intervals; compare to \cite{Takahashi2010} Fig.~14.}
\label{ccplot}
\end{figure}
\clearpage

>>>>

\cite{Takahashi2010} developed a mass density dataset using measurements from the GOES satellites over the years 1980 through 1992, with most measurements in the range $L=6.8\pm0.2$. The equatorial mass density, $\rho_{eq}$ was estimated from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$. The mass density $\rho$ was estimated using the the Alfven wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of an infinitely conducting ionosphere.

@@

In this work, events defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated in by horizontal dashed lines in the respective panels of Figure~\ref{AllData}.  In finding events, all fill values were replaced with linearly interpolated values.

@@

Figure~\ref{ccplot:a} shows the long-term trends of $\log(\rho_{eq})$ and $F_{10.7}$ computed using the median values in 27-day (non-overalpping?) windows using all non--fill hourly values.  A scatter plot of these two lines is shown in Figure~\ref{ccplot:b}.  The linear correlation is found to be $0.87$ (? plot says 0.94!).  In comparison, \cite{Takahashi2010} found a correlation coefficient of $0.94$ using measurements from all satellites in the time interval of 1980 through 1991.

\subsection{$D_{st}$ and $\rho_{eq}$ Events}

@@

In the upper panel of \ref{DailyAverages}, $D_{st}$ events are shown.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is $\sim 2$ amu/cm$^3$ instead of the value of $\sim 10$ amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for $D_{st}$, the number of measurements used in computing the medians is much larger as we use all available $D_{st}$ measurements instead of restricting to only values where a non--fill $\rho_{eq}$ value existed.

In the lower panel of \ref{DailyAverages}, $\rho_{eq}$ events are shown.  Here we see that $\rho_{eq}$ enhancements do not clearly correspond to a signature in $D_{st}$, but that there is a trend for such enhancments to occur when $B_{z}$ in the IMF is positive.  This is consistent with \cite{Takahashi2010} who note that very large increases in $\rho_{eq}$ often corresponded to movement of the plasmasphere to beyond geosynchronous distances.

This method finds 669 such periods between May 1983 and August 1991 with an average duration of 9 hours and a median duration of 3 hours. Figure \ref{Storms} shows the average values of all events over a window of 24 hours before onset and 48 hours after. Figure \ref{Storms}a shows events selected by looking for event onsets where $D_{st}$ crossed a threshold of $-40$~nT. The dashed red lines indicate plus and minus one standard deviation of values from all events that went into the average. The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack. A subset of longer duration events will be looked at later.

@@

In the upper panel of \ref{DailyAverages}, $D_{st}$ events are shown.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events corrspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is on the order of $\sim2$~amu/cm$^3$ instead of $\sim10$~amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for $D_{st}$, the number of measurements used in computing the medians is much larger as we use all available $D_{st}$ measurements instead of restricting to only values where a non-fill $\rho_{eq}$ value existed.

In the lower panel of \ref{DailyAverages}, $\rho_{eq}$ events are shown.  Here we see that $\rho_{eq}$ enhancements do not clearly correspond to a signature in $D_{st}$, but that there is a trend for such enhancments to occur when $B_{z}$ in the IMF is positive.  This is consistent with \cite{Takahashi2010} who note that very large increases in $\rho_{eq}$ often corresponded to movement of the plasmasphere to beyond geosynchronous distances.

@@

\section{More specific events}
It's hypothesized that progressively picking more specific event criteria will allow for the possibility of more significant results, at the expense of more bias in the selection process and potentially less overall usefulness of the results. That said, the predictability of extreme events is of definite interest, so an attempt has been made to find some reproducible method of prediction. Looking at events that last longer than 12 hours and events with an onset threshhold greater than $70$g/cm$^3$ results in the left and right sides of Figure \ref{Mspec} respectively.

@@

\begin{figure}[htp!]
\centering
\subcaptionbox{\label{ccplot:a}}
{\includegraphics[scale=0.40]{paperfigures/F107MDAllData.eps}}
\subcaptionbox{\label{ccplot:b}}
{\includegraphics[scale=0.40]{paperfigures/ccplot.eps}}
\caption{(a) 27-day averages of $F_{10.7}$ and $\rho_{eq}$. (b) Correlation between $\log(\rho_{eq})$ and $F_{10.7}$ using hour, day, and 27-day averaging intervals; compare to \cite{Takahashi2010} Fig.~14.}
\label{ccplot}
\end{figure}
\clearpage
